Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells

### 2024-06-28 Manuscript.tex
edits

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Interspecies comparisons of Mg/Ca ratios in limpet shells},
  pdfauthor={Niklas Hausmann; Donna Surge; Ivan Briz-Godino},
  pdfkeywords={Sclerochronology, Limpets, Elemental Ratio, Mg/Ca},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Interspecies comparisons of Mg/Ca ratios in limpet shells}
\author[1]{Niklas Hausmann%
\corref{cor1}%
}
 \ead{niklas@palaeo.eu} 
\author[2]{Donna Surge%
%
}

@@

\affiliation[1]{organization={Leibniz Zentrum für
Archäologie},addressline={Ludwig-Lindenschmit-Forum
1},city={Mainz},country={Germany},countrysep={,},postcode={55116},postcodesep={}}
\affiliation[2]{organization={University of North
Carolina},addressline={104 South Road, 225 Geology
Building},city={Chapel Hill,
NC},country={US},countrysep={,},postcode={27599-3315},postcodesep={}}
\affiliation[3]{organization={},country={Spain},countrysep={,},postcodesep={}}

@@

\begin{abstract}
This study provides a short reassessment of the use of Magnesium to
Calcium (Mg/Ca) ratios in Atlantic limpet shells to determine past sea
surface temperatures. While \emph{Patella vulgata} along the Spanish
shoreline has since then repeatedly produced reliable correlations
between sea surface temperature and Mg/Ca ratios, this is not the case
for other patelloid species. \emph{Patella vulgata} and \emph{Nacella
deaureata} have been studied using Mg/Ca with mixed or contrary results.
In this study, we present elemental maps of various such species
together with stable oxygen isotope values. Our dataset also includes
specimens that were previously unsuccessful in providing significant
correlations between \(\delta\)\textsuperscript{18}O and Mg/Ca ratios.
By reassessing these previous specimens and including a wider range of
modern and archaeological samples (n=12) from three patelloid species
(\emph{P. vulgata}, \emph{N. deaureata}, and \emph{N. magellanica}) we
further add to the growing set of evidence for the reliable use of Mg/Ca
ratios to detect palaeotemperature change and serve as a means to
determine ontogenetic age and season of capture as well as to reveal
locations of interest within the growth record (i.e.~annual temperature
minima and maxima) for the targeted analysis using
\(\delta\)\textsuperscript{18}O or clumped oxygen isotope analysis.
\end{abstract}

\begin{keyword}
    Sclerochronology \sep Limpets \sep Elemental Ratio \sep 
    Mg/Ca
\end{keyword}
\end{frontmatter}

@@

Limpet shells are commonly found within archaeological sites and and
past shorelines due to their robust carbonate structure and the
long-term use of limpets as a marine food source. They have successfully
been studied in the past in the context of coastal subsistence
economies, site occupation on seasonal
\citep{Shackleton1973-ij, Parker2018-wf, Bosch2018-ud} or long-term
scale \citep{Ortiz2015-mr}, as well as palaeotemperature
\citep{Fenger2007-gf, Surge2012-ba, Wang2012-ee, Colonese2012-ct, Ferguson2011-zl}.
Determining past sea surface temperature (SST) change largely relies on
the measurement of \(\delta\)\textsuperscript{18}O-values within the
calcitic parts of the shell, but attempts have been made to also use
elemental ratios, such as magnesium to calcium (Mg/Ca), to have an
alternative measure, that potentially provides a more accurate SST
estimate than \(\delta\)\textsuperscript{18}O-values, which are also
affected by changes in salinity. In addition, the data acquisition of
elemental ratios --- either through
laser-ablation-isotope-ratio-mass-spectrometry (LA-ICP-MS) or laser
induced breakdown spectroscopy (LIBS), can be much faster and
cost-effective, increasing the number of specimens that can be studied
overall \citep{Durham2017-fh, Hausmann2023-ih}.

Problematically, links between Mg/Ca ratios and SST changes in many
marine mollusc shells, including limpets, have shown to be unreliable
\citep{Surge2008-ri, Wanamaker2008-zl, Schone2010-yl, Freitas2012-tx, Graniero2015-zv, Poulain2015-dg, Vihtakari2017-wd}.
This is particularly the case where there is little available additional
information on metabolic processes, organic components of the shell
matrix, intra-increment and intra-shell variability, and growth rates,
which can independently and unpredictably affect Mg/Ca ratios and
confound their interpretation as temperature proxy. Confusingly,
multiple different temperature equations have been found for the same
species (see \citep{Freitas2012-tx, Vihtakari2017-wd} and references
therein). Coeval specimens sharing one locality can also show
differences in their relation to SST \citep{Hausmann2019-fi}. Where the
use of Mg/Ca as a palaeotemperature proxy was successful, anomalous
patterns in some specimens had still to be filtered out by hand,
reducing the overall robustness of the results of those successful
studies \citep{Ferguson2011-zl}.

While recent research particularly of \emph{Patella} sp. in the
Mediterranean and Southwest Europe have provided promising results
\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
There remains a lack of clarity for Atlantic limpet species,
particularly since it was last shown here, that they are not reliable
recorders of palaeotemperature \citep{Graniero2015-zv}.

In this study we will repeat and expand the analysis of Atlantic limpets
to determine the reliability of Mg/Ca ratios as palaeotemperature
proxies. To do this, we sampled a set of previously published and
unpublished limpet specimens dating to modern and archaeological
contexts using LIBS. LIBS allows us to carry out 2D imaging of entire
shell sections, which helps us to navigate the complex elemental
structure of the shell and better separate the external,
temperature-related changes from the internal and less understood
factors that influence the Mg/Ca ratio.

By relying and adding onto published datasets, we were able to
simultaneously avoid costs for new high-resolution
\(\delta\)\textsuperscript{18}O-data, to use real-world examples, and to
also provide pilot-data for areas of existing research interest.
Generally establishing the usefulness and reliability of Mg/Ca as SST
proxy in limpets should help to provide a platform for future research
and an important stepping stone to better understand elemental ratios in
other marine mollusc shells.

@@

The analysed specimens, their origin and respective studies with
research background information can be found in
Table\hyperref[Table_1]{1}. Here we will briefly summarise their
contextual information, which can be accessed in more detail at the
respective studies
\citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While those
studies also included other specimens, their accessibility or state of
preservation did not all lend them to be re-analysed.

@@

There are two main locations that we sourced specimens from: the Beagle
Channel in Tierra del Fuego (Argentina) and the island of West Ray in
Orkney (UK). Four specimens were collected from the Beagle Channel in
Cambaceres Bay in October 2016. The area receives around 570 mm
precipitation but the channel waters are predominantly influenced by
marine currents and in the period of October 2015 to October 2016 stayed
around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
August to 10ºC in February.

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
August 2009. Westray lies about 70 km north of the Scottish mainland and
experiences virtually no freshwater input with salinity values of 34.0
-- 34.5 psu \citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in
March to 13.8ºC in August (Figure~\ref{fig-SSTs}).

@@

In total, we studied 7 archaeological specimens of \emph{P. vulgata}
coming from Quoygrew on Westray dating to 900 -- 1200 CE. These shells
have previously been studied to characterise seasonal temperature change
during the Medieval Climate Anomaly and the Little Ice Age
\citep{Surge2012-ba}. Together they show a range of lifespans, with some
comparatively long for their size (QG1-7246-1, 12 years within 16 mm of
growth) making them particularly susceptible to time-averaging. That
said, the majority are under 5 years old. These shells are an ideal
comparative dataset from an archaeological assemblage to compare to the
modern shell ORK\_LT5 \citep{Graniero2017-io}.

@@

Mg/Ca elemental imaging was carried out using Laser Induced Breakdown
Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany)
using previously published methods \citep{Hausmann2023-ih}. These
involve the ablation of the shell section using an infrared (1064 nm)
laser (1--2 mJ, 100 Hz) onto a surface area of 20--30 µm to generate a
plasma plume. This plume emits light, which is measured using a
synchronised spectrometer. The resulting light spectrum quantifies the
emission lines of magnesium (MgII; 279.553 nm) and calcium (CaII;
315.887 nm) to determine their intensity ratio. While these two peaks
alone do not represent the molar concentrations of both elements (as
opposed to calibration free LIBS; e.g.
\citep{Martinez-Minchero2022-jz}), the intensity ratio is linearly
correlated to the molar concentration \citep{Hausmann2017-oa} and can be
used as a reliable indicator of Mg/Ca changes within the shell
carbonate.

Using this system we carried out elemental imaging of the shell
specimens at a resolution of 50 µm distance between sampling spots. Each
spot was irradiated 10 times with the 3 first spectra used for cleaning
steps and the remaining summed to get an average Mg/Ca intensity ratio
for each sample spot. Subsequently, we re-sampled the section using a
line scan at 10 µm resolution. This leads to an overlap between sample
spots, but also allows for a continuous record without gaps. Intensity
ratios were filtered for cases with high relative standard deviation
(i.e.~more than 10\%). These occurred in places where the previous
sampling procedures for carbonate powder as part of the oxygen isotope
analysis left an uneven sample surfaces introducing variability in the
plasma generation and thus uncertainty in the data of these locations.

@@

Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
limpet shells by species. Due to the high sampling resolution, all
shells show seasonally resolved and multi-year quasi-sinusoidal records
of SST change , with \emph{Nacella} sp. shells living 2--3 years and
\emph{Patella v}. regularly over 3 years. In those instances where
longer lifetimes are encountered (i.e.~specimen QG1-7246-1,
Figure~\ref{fig-Pat_iso} b).

@@

Interestingly, the annual ranges in
\(\delta\)\textsuperscript{18}O-values are not similar between the
coeval \emph{Nacella} sp. specimens with \(\delta\)\textsuperscript{18}O
values for the previous summer temperatures ranging from 0.4‰ to above
1.0‰, and for winter values between 2.7‰ and 2.0‰. \emph{P. vulgata} has
similarly variable \(\delta\)\textsuperscript{18}O-values for summer and
winter seasons, alas the lack of more that one modern specimens, means
that we cannot quantify the variability of recorded annual ranges
between specimens.

@@

Because the exact location used in the analysis of stable oxygen
isotopes is not possible to reconstruct for each carbonate powder
sample, we used dynamic time warping (DTW) to align the time series of
\(\delta\)\textsuperscript{18}O-values and Mg/Ca ratios. DTW is an
algorithm that measures similarity between two proxy-sequences, which
may vary in sampling resolution or interval. By stretching or
compressing sections of the series, DTW finds the probable alignment
between the two sequences. This allows us to compare the proxy data sets
more effectively, ensuring that the temporal dynamics of each shell are
accurately matched despite possible discrepancies in sampling intervals
or rates. We applied the DTW algorithm using the \texttt{dtw} package in
R \citep{Giorgino2009-sj, R_Core_Team2020-mk}, which provides a robust
framework for aligning time series data. This involved selecting
appropriate distance measures and constraints to ensure meaningful
alignment. The process involved iterative adjustments to minimise the
overall distance between corresponding points in the two data sets.

@@

The elemental imaging of \emph{P. vulgata} shells consistently showed
repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca
intensity ratios indicating and low ratios indicating low temperature
periods. As examples, Figure~\ref{fig-Pat_LIBS} shows the
2D-distribution of elemental ratios across the entire section of ORK-LT5
and the preserved anterior side of QG1-7188-1 (additional maps can be
found in the supplementaries). The repeating patterns are found across
the calcitic layers that are exterior to the myostracum (layers m+2 and
m+3) as well as the layers that are interior (m-2)
\citep{Fenger2007-gf}. The layer m-1, which is also interior, consists
of aragonite and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its
Mg/Ca intensity ratio consistently falls below the range of interest
(0.2 and above). Compared to other patelloid shells, whose interior is
almost entirely made of aragonite, this layer is very thin.

@@

Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
to its \(\delta\)\textsuperscript{18}O-values (similar graphs for other
specimen can be found in the supplementaries). The measurements of the
distance to the shell edge are not entirely identical, most likely
because one growth increment can have several different distances.
Depending on where one measures the distance and a line scan towards the
interior would be shorter than a line scan towards the exterior. That
said, increases and decreases of both proxies seem to mirror each other
and to be well aligned.

@@

Aligning the records programmatically using dynamic time warping,
allowed us to compare them directly to determine specimen-specific
equations and quantify the correlation of both proxies
(Figure~\ref{fig-Pat_Corr}). The equations for all specimens are
different but the majority (barring one: QG1-7246-1) do seem to cluster
around shared parameters (\(\delta\)\textsuperscript{18}Ο = 0.6 -- 0.065
* Mg/Ca). The various coefficients of determination
(R\textsuperscript{2}) range between 0.81 (QG1-7246-1) and 0.95
(QG1-7189-2). These specimens are also the oldest and youngest,
respectively, specimens, suggesting that lower growth rates and time
averaging had a negative effect on the correlation of both proxies. The
mean R\textsuperscript{2} value is 0.89, suggesting a generally good fit
between the two proxies.

@@

The LIBS data for \emph{Nacella} sp. specimens is less straightforward
with shells being much thinner and thus patterns being more difficult to
make out. Both specimens (ND-1016-3 and NM-1016-3) shown as example in
Figure~\ref{fig-Nac_LIBS} (see additional graphs in supplementaries)
also experience increases of Mg/Ca intensity ratio towards the exterior
of the shell, similar to the \emph{P. vulgata} specimens from Orkney.
This intra-increment heterogeneity combined with the thinness of the
shell, further complicates their analysis compared to \emph{P. vulgata}
or other \emph{Patella} species \citep{Hausmann2019-fi}. Nevertheless,
repeating patterns were visible in the anterior section of the shell,
which were also mirrored in the inner layers (m-2) of specimens
NM-1016-3 (Figure~\ref{fig-Nac_LIBS} c).

@@

\caption{\label{fig-Nac_LIBS}Example LIBS maps and line scans of
analysed specimens. (\emph{a+b}) ND-1016-3 and (\emph{c}+\emph{d})
NM-1016-3. Note that the ranges of Mg/Ca intensity ratios are not
identical but are chosen arbitrarily to increase contrast. Areas in grey
consist of low-Mg aragonite and fall outside the chosen colour-range.
Note the inset in (\emph{c}) where the shell edge has been resampled
using a higher resolution (50 µm instead of 100 µm) to better understand
the Mg/Ca ratios at the shell edge.}

@@

Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} sp.
shells were much lower (0.05--0.30) than those seen in other patelloid
species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in )
using the same emission lines. This is particularly visible in
Figure~\ref{fig-Nac_LIBS} c, where Mg/Ca intensity ratios range from
\textasciitilde0.5--0.7 in \citep{Hausmann2023-ih}). While the intensity
ratio depends chiefly on the chosen emission lines to calculate the
proportion of the chosen Magnesium to the Calcium peak, the low ratios
still indicate lower concentrations than seen elsewhere, or seen only in
aragonitic parts of e.g.~\emph{Patella} shells. Similar observations
have been made before \citep{Graniero2017-io}.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic limpet shells},
  pdfauthor={Niklas Hausmann; Donna Surge; Ivan Briz-Godino},
  pdfkeywords={Sclerochronology, Patella, Nacella, LIBS, Mg/Ca},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Confirmation of Mg/Ca ratios as palaeothermometers in Atlantic
limpet shells}
\author[1]{Niklas Hausmann%
\corref{cor1}%
}
 \ead{niklas@palaeo.eu} 
\author[2]{Donna Surge%
%
}

@@

\affiliation[1]{organization={Leibniz Zentrum für
Archäologie},addressline={Ludwig-Lindenschmit-Forum
1},city={Mainz},country={Germany},countrysep={,},postcode={55116},postcodesep={}}
\affiliation[2]{organization={University of North
Carolina},addressline={104 South Road, 225 Geology
Building},city={Chapel Hill,
NC},country={US},countrysep={,},postcode={27599-3315},postcodesep={}}
\affiliation[3]{organization={Department of Evolutionary Biology,
Ecology and Environmental Science, Biodiversity Research Institute
(IRBio), University of Barcelona},addressline={Avinguda Diagonal,
643},city={Barcelona},country={Spain},countrysep={,},postcode={8028},postcodesep={}}

@@

\begin{abstract}
This study provides a reassessment of the use of Magnesium to Calcium
(Mg/Ca) ratios in Atlantic limpet shells to determine past sea surface
temperatures (SST). While \emph{Patella depressa} along the Spanish
shoreline has repeatedly produced reliable correlations between SST and
Mg/Ca ratios, this is not the case for other patelloid species.
\emph{Patella vulgata}, \emph{Nacella deaureata}, and \emph{Nacella
magellanica} have been studied using Mg/Ca with mixed or contrary
results. In this study, we present elemental maps of various such
species together with stable oxygen isotope values that confirm a good
relationship with SST. Our dataset also reassesses a specimen which was
previously unsuccessful in providing significant correlations between
\(\delta\)\textsuperscript{18}O and Mg/Ca ratios. By reassessing these
species and including modern and archaeological samples (n=12) from
three patelloid species (\emph{P. vulgata}, \emph{N. deaureata}, and
\emph{N. magellanica}) we further add to the growing set of evidence for
the reliable use of Mg/Ca ratios to detect palaeotemperature change. As
a result, these species can in the future serve as a means to determine
ontogenetic age and season of capture as well as to reveal locations of
interest within the growth record (i.e.~annual temperature minima and
maxima) for the targeted analysis using \(\delta\)\textsuperscript{18}O
or clumped oxygen isotope analysis.
\end{abstract}

\begin{keyword}
    Sclerochronology \sep Patella \sep Nacella \sep LIBS \sep 
    Mg/Ca
\end{keyword}
\end{frontmatter}

@@

Limpet shells are commonly found within archaeological sites and and
past shorelines due to their robust carbonate structure and the
long-term use of limpets as a marine food source \citep{Robson2023-oz}.
They have successfully been studied in the past in the context of
coastal subsistence economies, site occupation on seasonal
\citep{Shackleton1973-ij, Parker2018-wf, Bosch2018-ud} or long-term
scale \citep{Ortiz2015-mr}, as well as palaeotemperature
\citep{Fenger2007-gf, Surge2012-ba, Wang2012-ee, Colonese2012-ct, Ferguson2011-zl}.
Determining past sea surface temperature (SST) change largely relies on
the measurement of \(\delta\)\textsuperscript{18}O-values within the
calcitic parts of the shell, but attempts have been made to also use
elemental ratios, such as magnesium to calcium (Mg/Ca)
\citep{Graniero2017-io, Ferguson2011-zl}, to have an alternative
measure, that potentially provides a more accurate SST estimate than
\(\delta\)\textsuperscript{18}O-values, which are also affected by
changes in salinity. In addition, the data acquisition of elemental
ratios --- either through laser-ablation-isotope-ratio-mass-spectrometry
(LA-ICP-MS) or laser induced breakdown spectroscopy (LIBS), can be much
faster and cost-effective, increasing the number of specimens that can
be studied overall \citep{Durham2017-fh, Hausmann2023-ih}.

Problematically, links between Mg/Ca ratios and SST changes in many
marine mollusc shells, including limpets, have shown to be unreliable
\citep{Surge2008-ri, Wanamaker2008-zl, Schone2010-yl, Freitas2012-tx, Graniero2017-io, Poulain2015-dg, Vihtakari2017-wd}.
This is particularly the case where there is little available additional
information on metabolic processes, organic components of the shell
matrix, intra-increment and intra-shell variability, and growth rates,
which can independently and unpredictably affect Mg/Ca ratios and
confound their interpretation as temperature proxy. Confusingly, in some
cases multiple different temperature equations have been found for the
same species (see Freitas \emph{et al}. \citeyearpar{Freitas2012-tx} and
Vihtakari \emph{et al}. \citeyearpar{Vihtakari2017-wd} and references
therein). Coeval specimens sharing one locality can also show
differences in their relation to SST \citep{Hausmann2019-fi}. Where the
use of Mg/Ca as a palaeotemperature proxy was successful, anomalous
patterns in some specimens had still to be filtered out by hand,
reducing the overall robustness of the results of those successful
studies \citep{Ferguson2011-zl}.

While recent research particularly of \emph{Patella} sp. in the
Mediterranean and Southwest Europe have provided promising results
\citep{Hausmann2019-fi, Garcia-Escarzaga2015-jc, Garcia-Escarzaga2018-nf}.
There remains a lack of clarity for Atlantic limpet species,
particularly since it was last shown here, that they are not reliable
recorders of palaeotemperature \citep{Graniero2017-io}.

In this study we will repeat and expand the analysis of Atlantic limpets
to determine the reliability of Mg/Ca ratios as palaeotemperature
proxies. To do this, we sampled a set of previously published and
unpublished limpet specimens dating to modern and archaeological
contexts using LIBS. LIBS allows us to carry out 2D imaging of entire
shell sections, which helps us to navigate the complex elemental
structure of the shell and better separate the external,
temperature-related changes from the internal and less understood
factors that influence the Mg/Ca ratio. By working with already
published datasets, we were able to simultaneously avoid costs for new
high-resolution \(\delta\)\textsuperscript{18}O-data, to use real-world
examples, and to also provide pilot-data for areas of existing research
interest. Generally establishing the usefulness and reliability of Mg/Ca
as SST proxy in limpets should help to provide a platform for such
future research and an important stepping stone to a better
understanding of elemental ratios in other marine mollusc shells.

@@

The analysed specimens, their origin and respective studies with
research background information can be found in Table
\hyperref[Table_1]{1}. Here we will briefly summarise their contextual
information, which can be accessed in more detail at the respective
studies \citep{Nicastro2020-ih, Surge2012-ba, Graniero2017-io}. While
those studies also included other specimens, their accessibility or
state of preservation did not allow them to be re-analysed.

@@

There are two main locations that we sourced specimens from: the Beagle
Channel in Tierra del Fuego (Argentina) and the island of West Ray in
Orkney (UK). Four modern specimens were collected from the Beagle
Channel in Cambaceres Bay in October 2016. The area receives around 570
mm precipitation but the channel waters are predominantly influenced by
marine currents and in the period of October 2015 to October 2016 stayed
around 30.7 ± 0.7 psu \citep{Nicastro2020-ih}. SSTs range from 4.7ºC in
August to 10ºC in February (Figure~\ref{fig-SSTs} - red).

On Westray one modern shell (ORK-LT5) was collected in Rack Wick Bay in
August 2009 and has previously been analysed using stable carbon and
oxygen isotopes, as well as LA-ICP-MS \citep{Graniero2017-io}. Westray
lies about 70 km north of the Scottish mainland and experiences
virtually no freshwater input with salinity values of 34.0 -- 34.5 psu
\citep{Inall2009-ho}. The annual SST ranges from 6.3ºC in March to
13.8ºC in August (Figure~\ref{fig-SSTs} - blue).

@@

We also studied 7 archaeological specimens of \emph{P. vulgata} coming
from Quoygrew on Westray dating to 900 -- 1200 CE. These shells have
previously been studied to characterise seasonal temperature change
during the Medieval Climate Anomaly and the Little Ice Age
\citep{Surge2012-ba}. Together they show a range of lifespans, with some
comparatively long for their size (QG1-7246-1, 12 years within 16 mm of
growth) making them particularly susceptible to time-averaging. That
said, the majority are under 5 years old.

@@

Mg/Ca elemental imaging was carried out using Laser Induced Breakdown
Spectroscopy at the Leibniz Zentrum für Archäologie (Mainz-Germany)
using previously published methods \citep{Hausmann2023-ih}. These
involve the ablation of the shell section using an infrared (1064 nm)
laser (1--2 mJ, 100 Hz) onto a surface area of 20--30 µm to generate a
plasma plume. This plume emits light, which is measured using a
synchronised spectrometer (delay time = 0.5 µs, integration time = 1
µs). The resulting light spectrum quantifies the emission lines of
Magnesium (MgII; 279.553 nm) and Calcium (CaII; 315.887 nm) to determine
their intensity ratio. While these two peaks alone do not represent the
molar concentrations of both elements (as opposed to calibration free
LIBS; e.g. \citep{Martinez-Minchero2022-jz}), the intensity ratio is
linearly correlated to the molar concentration \citep{Hausmann2017-oa}
and can be used as a reliable indicator of Mg/Ca changes within the
shell carbonate. All values reported here at in arbitrary units based on
the ratio of intensities of the peaks above.

Using this system we carried out elemental imaging of the shell
specimens at a resolution of 50--100 µm distance between sampling spots.
Each spot was irradiated 10 times with the 3 first spectra discarded as
a cleaning step and the remaining spectra summed to get an average Mg/Ca
intensity ratio for each sample spot. Subsequently, we re-sampled the
section using a line scan at 10 µm resolution. This leads to an overlap
between sample spots, but also allows for a continuous record without
gaps. Intensity ratios were filtered for cases with high relative
standard deviation (i.e.~more than 10\%). These occurred in places where
the previous sampling procedures for carbonate powder as part of the
oxygen isotope analysis left an uneven sample surface introducing
variability in the plasma generation and thus uncertainty in the data of
these locations.

@@

Figure~\ref{fig-Nac_iso} and Figure~\ref{fig-Pat_iso} show the
previously acquired sequential \(\delta\)\textsuperscript{18}O-values of
limpet shells by species (\emph{Nacella} sp., and \emph{Patella
vulgata}, respectively. Due to the high sampling resolution, all shells
show seasonally resolved and multi-year quasi-sinusoidal records of SST
change, with \emph{Nacella} sp. shells living 2--3 years and \emph{P.
vulgata} living regularly over 3 years. In those instances where longer
lifetimes are encountered (i.e.~specimen QG1-7246-1,
Figure~\ref{fig-Pat_iso} b) the temporal-resolution is reduced from more
than a dozen of samples per year, to only around 7.

@@

@@

Because it is not possible to directly associate the exact sample
locations used in the analysis of stable oxygen isotopes with the LIBS
data, we used dynamic time warping (DTW) to align the time series of
\(\delta\)\textsuperscript{18}O-values and Mg/Ca ratios. DTW is an
algorithm that measures similarity between two proxy-sequences, which
may vary in sampling resolution or interval. By stretching or
compressing sections of the series, DTW finds the probable alignment
between the two sequences. This allows us to compare the proxy data sets
more effectively, ensuring that the temporal dynamics of each shell are
accurately matched despite possible discrepancies in sampling intervals
or rates. We applied the DTW algorithm using the \texttt{dtw} package in
R \citep{Giorgino2009-sj, R_Core_Team2020-mk}, which provides a robust
framework for aligning time series data. This involved selecting
appropriate distance measures and constraints to ensure meaningful
alignment. The process involved iterative adjustments to minimise the
overall distance between corresponding points in the two data sets.

@@

The elemental imaging of \emph{P. vulgata} shells consistently showed
repeating patterns of Mg/Ca intensity ratio changes with high Mg/Ca
intensity ratios indicating and low ratios indicating low temperature
periods. As examples, Figure~\ref{fig-Pat_LIBS} shows the
2D-distribution of elemental ratios across the entire section of ORK-LT5
and the preserved anterior side of QG1-7188-1 (additional maps can be
found in the supplementaries). The repeating patterns are found across
the calcitic layers that are exterior to the myostracum (layers m+2 and
m+3) as well as the layers that are interior (m-2)
\citep{Fenger2007-gf}. The layer m-1, which is also interior, consists
of aragonite and is seen in Figure~\ref{fig-Pat_LIBS} in grey, as its
Mg/Ca intensity ratio consistently falls below the range of interest
(0.2 and above). Compared to other patelloid shells, some of whose
interior is almost entirely made of aragonite, this layer is very thin.

@@

Figure~\ref{fig-Pat_Comp} shows the line scan of ORK-LT5 in comparison
to its \(\delta\)\textsuperscript{18}O-values (similar graphs for other
specimen can be found in the supplementaries). The measurements of the
distance to the shell edge are not entirely identical, most likely
because one growth increment can have a range of distances to the edge,
depending on where one measures, and a line scan along the interior
would be shorter than a line scan along the exterior. That said,
increases and decreases of both proxies seem to mirror each other and to
be well aligned.

@@

Aligning the records programmatically using dynamic time warping,
allowed us to compare them directly to determine specimen-specific
equations and quantify the correlation of both proxies
(Figure~\ref{fig-Pat_Corr}). The equations for all specimens are
different but the majority (barring one: QG1-7246-1) do seem to cluster
around shared parameters (\(\delta\)\textsuperscript{18}Ο = 0.6 -- 0.065
* Mg/Ca). The various coefficients of determination
(R\textsuperscript{2}) range between 0.81 (QG1-7246-1) and 0.95
(QG1-7189-2). These specimens are also the oldest and youngest
specimens, respectively, suggesting that lower growth rates and time
averaging had a negative effect on the correlation of both proxies. The
mean R\textsuperscript{2} value of the individual \emph{P. vulgata}
specimens is 0.89, suggesting a generally good fit between the two
proxies.

@@

The LIBS data for \emph{Nacella} sp. specimens is less straightforward
with shells being much thinner and thus patterns being more difficult to
make out. Both specimens (ND-1016-3 and NM-1016-3) shown as example in
Figure~\ref{fig-Nac_LIBS} (see additional graphs in supplementaries)
also experience increases of Mg/Ca intensity ratio towards the exterior
of the shell, similar to the \emph{P. vulgata} specimens from Orkney.
This intra-increment heterogeneity combined with the thinness of the
shell, further complicates their analysis compared to \emph{P. vulgata}
or other \emph{Patella} species \citep{Hausmann2019-fi}. Nevertheless,
repeating patterns were visible in the anterior section of the shell,
which were also mirrored in the inner layers (m-2) of specimens
NM-1016-3 (Figure~\ref{fig-Nac_LIBS} b).

@@

\caption{\label{fig-Nac_LIBS}Example LIBS maps and line scans of
analysed \emph{Nacella} sp. specimens. (\emph{a+c}) ND-1016-3 and
(\emph{b}+\emph{d}) NM-1016-3. Note that the ranges of Mg/Ca intensity
ratios are not identical but are chosen arbitrarily to increase
contrast. Areas in grey consist of low-Mg aragonite and fall outside the
chosen colour-range. Note the inset in (\emph{c}) where the shell edge
has been resampled using a higher resolution (50 µm instead of 100 µm)
to better understand the Mg/Ca ratios at the shell edge.}

@@

Interestingly, the Mg/Ca intensity ratios in the \emph{Nacella} sp.
shells were much lower (0.05--0.30) than those seen in other patelloid
species (e.g.~0.3--0.7 in \emph{Patella vulgata} above, or 0.5--1.5 in
Hausmann \emph{et al}. \citeyearpar{Hausmann2023-ih}) using the same
emission lines. While the intensity ratio depends chiefly on the chosen
emission lines to calculate the proportion of the chosen Magnesium to
the chosen Calcium peak, the low ratios still indicate lower
concentrations than seen elsewhere, or seen only in aragonitic parts of
e.g.~\emph{Patella} shells. Similar observations have been made before
\citep{Graniero2017-io}.
